---
title: "Interactive Data Visualization with Plotly"
author: "Gabriel Demetrios Lafis"
date: "May 28, 2025"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(dplyr)
```

## Introduction

This presentation demonstrates the power of interactive data visualization using Plotly in R.

- Plotly is a powerful visualization library that creates interactive web-based graphs
- Interactive visualizations allow users to explore data more deeply
- Perfect for presentations, dashboards, and data exploration

## The Dataset: Motor Trend Car Road Tests

We'll be using the famous `mtcars` dataset:

```{r}
# Load the data
cars <- read.csv("data/mtcars.csv")
# Display the first few rows
head(cars, 3)
```

This dataset includes data from 32 automobiles from Motor Trend magazine (1973-1974).

## Cars by Number of Forward Gears and Cylinders

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Create a grouped bar chart showing the distribution of cars by cylinders and gears
p <- cars %>%
  count(cyl, gear) %>%
  plot_ly(x = ~cyl, y = ~n, color = ~factor(gear), type = "bar") %>%
  layout(
    title = "Distribution of Cars by Cylinders and Gears",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Count"),
    barmode = "stack",
    legend = list(title = "Number of Gears")
  )

p
```

## Enhanced Visualization: Horsepower vs. MPG

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Create an interactive scatter plot
p2 <- plot_ly(cars, x = ~hp, y = ~mpg, color = ~factor(cyl),
             size = ~wt, sizes = c(20, 150),
             text = ~paste("Car: ", model, 
                          "<br>MPG: ", mpg,
                          "<br>Horsepower: ", hp,
                          "<br>Weight: ", wt,
                          "<br>Cylinders: ", cyl),
             hoverinfo = "text") %>%
  layout(
    title = "Horsepower vs. Fuel Efficiency",
    xaxis = list(title = "Horsepower"),
    yaxis = list(title = "Miles Per Gallon"),
    legend = list(title = "Cylinders")
  )

p2
```

## Relationship Between Weight and Quarter Mile Time

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Create an interactive scatter plot with trend line
p3 <- plot_ly(cars, x = ~wt, y = ~qsec) %>%
  add_markers(
    color = ~factor(am), 
    colors = c("#FF5733", "#33A2FF"),
    symbol = ~factor(am),
    symbols = c("circle", "diamond"),
    size = ~hp/10,
    text = ~paste("Car: ", model, 
                 "<br>Weight: ", wt, " tons",
                 "<br>Quarter Mile: ", qsec, " sec",
                 "<br>Transmission: ", ifelse(am == 0, "Automatic", "Manual")),
    hoverinfo = "text"
  ) %>%
  add_lines(
    data = cars,
    x = ~wt,
    y = fitted(lm(qsec ~ wt, data = cars)),
    line = list(color = 'rgba(0, 0, 0, 0.5)', width = 2),
    showlegend = FALSE,
    name = "Trend Line"
  ) %>%
  layout(
    title = "Car Weight vs. Quarter Mile Time",
    xaxis = list(title = "Weight (tons)"),
    yaxis = list(title = "Quarter Mile Time (sec)"),
    legend = list(title = "Transmission",
                 labels = c("Automatic", "Manual"))
  )

p3
```

## Conclusion

### Benefits of Interactive Visualizations

- **Engagement**: Users can explore data on their own terms
- **Depth**: Multiple dimensions of data can be represented simultaneously
- **Clarity**: Complex relationships become more apparent
- **Accessibility**: Technical and non-technical audiences can understand insights

### Next Steps

- Explore more advanced Plotly features
- Integrate with Shiny for even more interactivity
- Create comprehensive dashboards for data analysis
